%\begin{abstract}
%Neutron stars are the best.
%\end{abstract}


%When neutron stars (NS) accrete gas from low-mass binary companions, explosive nuclear burning reactions in the NS envelope fuse hydrogen and helium into heavier elements. The resulting thermonuclear (type-I) X-ray bursts produce energy spectra that are fit well with black bodies, but a significant number of burst observations show deviations from Planck spectra.
%
%Neutron star (NS) masses and radii can be estimated from observations of photospheric radius- expansion X-ray bursts, provided the chemical composition of the photosphere, the spectral colour-correction factors in the observed luminosity range, and the emission area during the bursts are known. 
%
%The cooling phase of thermonuclear (type-I) X-ray bursts can be used to constrain neutron star (NS) compactness by comparing the observed cooling tracks of bursts to accurate theoretical atmosphere model calculations. 


%--------------------------------------------------
%intro-intro
%First, a short history of neutron stars is presented, focusing on both theoretical and observational results.
%After this, an introduction to the basic physics of stars is given.
%intro-eos
%Next, the physics of neutron star interiors is discussed in more detail.
%The discussion first focuses on the uppermost layers of the star called an atmosphere.
%This thin layer of plasma is responsible of shaping the emergent radiation.
%Below the atmosphere is a solid layer called the crust.
%In order to understand the physics of the crust, an introduction to the degenerate matter is presented.
%The matter in the crust can then be described by the degenerate electron gas model.
%The crust is, however, only couple of kilometers thick and a bulk of the star consists of a liquid core.
%The EoS of the matter in the core is largely unknown and because of this, we present a way to parameterize the nuclear physics using a so-called polytropic model.
%intro-astro
%The introduction is continued by describing the colorful astrophysics of neutron stars.
%We mainly focus on presenting the basics of accretion, an astrophysical process where matter is transferred into a compact object such as a neutron star.
%intro-constraints
%Lastly, we show how the neutron star size can be inferred from the X-ray burst observations.



\vspace{2.0cm}
\chapter*{Abstract}

%background
Neutron stars are one of the most dense objects in the Universe.
However, the exact description of the equation of state (EoS) of the cold ultra-dense matter inside them is still a mystery.
In this thesis, we measure the size of some neutron stars using astrophysical observations of so-called X-ray bursts that are produced by thermonuclear runaways in the uppermost layers of the star.
By measuring the size, we can then set constraints on the nuclear physics of the unknown interiors and ultimately on the EoS of the cold dense matter.

The size measurements are done by comparing the cooling of the neutron star surfaces after the bursts to theoretical atmosphere model calculations.
Hence, accurate modeling of the emergent radiation from the atmospheres is needed.
In the first part of this thesis, I have studied how the emergent spectra differ if the atmosphere is enriched with nuclear burning ashes from the bursts.
This gives us new tools to understand and interpret the X-ray burst observations.
In addition, I have shown how the emerging radiation is modified when it originates from rapidly rotating oblate neutron stars.

Together with accurate atmosphere models, we must also be careful in selecting only those bursts that are not actively heated by the infalling material.
In the second part of the thesis, I have focused on studying the astrophysical environments of the X-ray bursts in order to quantify the effect of accretion on the mass and radius measurements.
Importantly, it is shown that only the bursts that occur during the low accretion rate state can be used for the size determination because otherwise the accretion flow might influence the cooling of the star's surface.

After taking these steps into account, it is possible to put constraints on the mass, radius, distance, and atmosphere composition of neutron stars exhibiting X-ray bursts.
In the third part of the thesis, I have used the aforementioned models and methods to constrain the mass and radius of neutron stars using the low accretion rate X-ray bursts.
The method has been applied to three neutrons stars in low mass X-ray binary systems 4U 1702$-$429, 4U 1724$-$307, and SAX J1810.8$-$260 for which the radius is measured to be between $10.5 - 12.8\km$ ($95$\% credibility).
The newly computed atmosphere models have also been used to detect a presence of burning ashes in the atmosphere of the neutron star in HETE J1900.1$-$2455.
Later on, an improved Bayesian method of fitting the atmosphere models directly to the observed spectra has also improved the radius constraints of 4U 1702$-$429 to $R = 12.4 \pm 0.4\km$ ($68\%$ credibility). 
These results are in a good agreement with the current nuclear physical predictions and showcase how astrophysical measurements can be used to gauge the unknown nuclear physics of neutron stars.

%method results
%conclusions



\vspace{2.0cm}
\chapter*{Tiivistelmä}

Neutronitähdet ovat yksiä Universumimme tiheimpiä kappaleita.
Niiden sisältämän ultratiheän kylmän aineen tilanyhtälö ja tarkka käyttäytyminen on kuitenkin vielä tuntematonta.
Tässä väitöskirjassa näytän kuinka kaukaisenkin neutronitähden koko voidaan mitata käyttäen niin kutsuttujen röntgenpurkausten lähettämää säteilyä.
Röntenpurkaukset itsessään saavat alkunsa termisestä fuusioreaktiosta joka tuottaa valtaisan räjähdyksen tähden pintakerroksissa.
Mittaamalla ja mallintamalla näistä purkauksista lähetävää säteilyä, voimme myös lopulta sanoa jotain neutronitähtien sisältämän aineen käyttäytymisestä ja siten myös kylmän tiheän aineen tilanyhtälöstä.

Mittaukset tehdään vertaamalla neutronitähtien pinnalta alkunsa saavaa säteilyä teoreettisiin ilmakehämalleihin jotka ennustavat kuinka pinnan tulisi jäähtyä purkausten jälkeen.
Tämän takia tarvitsemme tarkkoja malleja säteilyn kulusta ilmakehässä.
Ensimmäisessä osassa väitöskirjaani olen tutkinut kuinka ilmakehässä olevat raskaammat fuusioreaktioissa syntyneet alkuaineet vaikuttavat tämän säteilyn etenemiseen ilmakehän plasmassa.
Tämä auttaa meitä ymmärtämään ja tulkitsemaan myös itse röntgenpurkausten havaintoja.
Lisäksi olen näyttänyt kuinka havaittu säteily voi muuttua yleisen suhteellisuusteorian ansiosta koska säteilevät neutronitähdet ovat yleensä erittäin nopeasti pyöriviä ja navoilta litistyneitä.

Tarkkojen ilmakehämallien lisäksi meidän täytyy myös ymmärtää mitä neutronitähden ympärillä tapahtuu.
Väitöskirjani toisessa osassa tutkin kuinka tämä ympäristö voi vaikuttaa herkkiin tähden säteen mittauksiin, sillä joskus neutronitähden pinnalle putoava materia voi hankaloittaa mittauksia.
Tärkeimpänä löydöksenä olemme näyttäneet että ainoastaan purkauksia jotka tapahtuvat, kun putoaavaa materiaa on erittäin vähän, voidaan luotettavasti käyttää säteen mittaamiseen.

Otammalla nämä seikat huomioon, on sen jälkeen mahdollista mitata neutronitähden koko, etäisyys, ja ilmakehän koostumus vertaamalla oikeiden, havaittujen röntgenpurkausten jäähtymistä mallien ennusteisiin.
Viimeisessä osassa väitöskirjaani olen tutkinut röntenpurkausten säteilyä kolmesta eri neutronitähdesta jotka sijaitsevat kaksoistähtijärjestelmissä 4U 1702$-$429, 4U 1724$-$307, ja SAX J1810.8$-$260.
Näiden säteen olen mitannut olevan $10.5$ ja $12.8$ km välillä ($95\%$ luottamustaso).
Olemme myös käyttäneet uusia ilmakehämallejani havaitsemaan ja todistamaan kuinka kaksoistähtijärjestelmässä HETE J1900.1$-$2455 sijaitsevan neutronitähden pintakerrokset sisältävät raskaita alkuaineita jotka ovat syntyneet fuusioreaktion aikana.
Onnistuin myös uudenlaisella parannellulla Bayesilaisella metodilla, missä ilmakehämalleja sovitetaan suoraan röntgenhavaintoihin, mittaamaan 4U 1724$-$429:ssä sijaitsevan neutronitähden säteeksi $R=12.4 \pm 0.4\km$ ($68\%$ luottamustaso).
Nämä uudet tulokset ovat hyvin sopusoinnussa uusien ydinfysiikasta tulevien ennusteiden kanssa ja lisäksi näyttävät kuinka astrofysikaalisia mittauksia voidaan käyttää apuna ydinfysiikan tutkimuksessa.

