\chapter*{Acknowledgements}

%Looking back on what I actually did during my Ph.D. studies it seems that I applied some kind of a Monte Carlo random algorithm on what I selected as my focus of the year.
%On my first year of studies, I learned to program.
%On my second year, I learned basic statistics and Bayesian inference.
%During my third year, I learned mathematics.
%On my last, fourth year I finally grasped some simple astrophysics.
%In the last couple of months when starting this thesis project I finally realized what it is about.
%During the last days of writing this thesis, I learned not to care too much about non-sense details and just to get it done.
%It seems that exactly the opposite order of learning these things would have been optimal.
\vspace{-0.9cm}
This has been quite a journey and now when thinking about it, I have no idea how I even ended up here.
When I started as an undergraduate in University of Oulu, I studied biophysics and theoretical physics.
For some unknown reason I still ended up doing my Bachelor's thesis in astrophysics.
The topic was X-ray bursts from neutron stars.
Now, couple of years later, I am finishing up a PhD thesis on X-ray bursts from neutron stars.
What started as a short Bachelor's project turned out to be quite a colorful academic experience for me.
Hence, multiple thanks are also due for many different people.

%After that, I quickly followed by a Master's thesis project on ``Neutron star mass and radius constraints from X-ray bursts''. 
%Four years later I am now finishing up a Ph.D. thesis on almost the same topic.
%After reading that Master's thesis now, it seems that I had no clue about this stuff.
%Maybe same applies to this thesis after 5 years.

%#Supervisors:
%Juri
%Jari
%Sergei

Firstly, I would like to thank my supervisors Prof. Juri Poutanen, Dr. Jari Kajava, and Dr. Sergei Tsygankov.
This whole project would not have been possible without the help, guidance, and mentoring that I received from you.

%money
%I acknowledge the support from Väisälä Foundation and University of Turku Doctoral Programme in Physical and Chemical Sciences.

%#Defence:
%Ed Brown
%Jean in't Zand
%Fred Lamb
 
I would also like to thank Prof. Frederick K. Lamb from University of Illinois at Urbana-Champaign and Dr. Jean in 't Zand from SRON Netherlands Institute for Space Research for their extremely careful pre-examination of this thesis.
All the remaining errors in this thesis are mine.
I would also like to thank Prof. Edward Brown from Michigan State University for agreeing to become my esteemed opponent.

%#Other scientific ppl:
%Andrew
%Cole
%Valery
%Sasha
%Pauli
%Aleksi Vuorinen
 
Scientific work is not done in a vacuum.
Hence, warm thanks should also be given to my collaborators.
I would like to thank Andrew Steiner, Cole Miller, Valery Suleimanov, Aleksi Vuorinen, and Eemeli Annala, to name a few.
It has been a pleasure to work with you!
 
%#group:
%Tuomo
%Jere
%Ilya
%Anna
%Pavel
%Vlad

%#HPC:
%Pekka M.
%Jussi E.
%Martti L.
%Mikko B.
%Sami I.
%Sebastian vA.
%Seija 

%#Observatory:
%Robertto
%Ronald
 
Warm thanks also for the members and previous members of our high-energy astrophysics research group that created a nice and welcoming atmosphere in Tuorla and in Oulu for me.
These include (but is not limited to): Sasha Veledina, Sasha Mushtukov, Pavel, Tuulia, Anna, Ilia, Jere, Juhani, Teppo, and Tuomo.
Thank you also for the other members of the Tuorla Observatory, I had fun time staying and working in Tuorla with you.
Special thanks to Sirpa Reinikainen and Sanna Ranto for helping me fight against bureaucracy.
Warm thanks also to my friends Roberto for being punctually at our office door at 12:30 every day for lunch and Ronald for all the sauna/music/bar/etc. sessions.
Big thanks should also be given to the people at CSC who I had a pleasure to work with: 
thanks Jussi, Martti, Mikko, Sami, Sebastian, Seija, and Pekka. 
It has been a pleasure to work (and to continue working) with you!
 
 
%#Hese:
%Jaakko
%Joonas I.
%Joonas S.
%Joonas M.
%Miku
%Vesku
%Harri
%Olli
 
Some life exists outside work too. 
Mainly in Helsinki.
Warm thanks to my friends Joonas, Joonas, Joonas, Jaakko, Miku, Vesku, Harri, Olli, Maija, Suvi, Janne, Henri, and Pauli, to name a few.


%#perhe:
%Ämmi & Eki
%Mummi & Ukki
%
%Vilma
%Severi
%
%Äiti
%Isä

Lämpimät kiitokset myös perheelleni kaikesta tuesta.
Kiitos Ämmi ja Eki sekä Mummi ja Ukki. 
Kiitos Severi ja Vilma.
Kiitos Äiti ja Isä.
Ilman teitä en olisi tässä.

%Venla
Finally and most importantly, my deepest gratitude goes to Venla for actually making this (and many other things) possible. I could not have done this without you.
